Global adjoint tomography -- Model GLAD-M25\footnote{This paper is dedicated to Dimitri Komatitsch, computational geophysicist extraordinaire, who passed away unexpectedly shortly before its submission.}

### 2020-02-19 response_to_reviewers.tex

\rev{The manuscript is in decent shape but the writing is somewhat rushed after page 20 or so. The senior authors should polish the text and remove paragraphs, primarily in early sections, on topics that have been covered by Bozdag et al.
The authors don't do the point-spread function analysis so subsection 9.1 can be eliminated. I have annotated the PDF copy to suggest edits and to point out fuzzy wording.
}

\response{We have further polished the writing later in the manuscript
Thanks for the annotated manuscript! We have accommodated most comments and suggestions.
Based on suggestions by the other reviewers,
we added six PSF tests to the manuscript and the supplementary material.}

\rev{The GLAD-M tomography stands out from the other approaches by casting the inverse without compromising the physics of wave propagation. Consequently, the workflow is laborious and iterations are slow. Ten iterations for a moderate data set (compared to the S40RTS data set, for example), took more than two years. Sections 1--9 indicate that the authors are convinced about the value of their approach but the discussion onward from Section 10 is qualitative and not serving the authors' case.
I worry that readers might wonder whether M25 is a significant step forward from M15 and whether the tremendous engineering and use of computer power justifies the (re)discovery of velocity anomalies that do not appear to change our understanding of mantle structure and dynamics.
}

\response{We have modified the manuscript substantially, as discussed previously. We focus more concretely on differences between M15 and M25 to better highlight the power of 10 additional iterations with $>1,400$ events. The ``two-years'' is not just the time spent for 10 iterations but code development and optimization as well as investigations of global FWI to define strategies for how to make progress. This is not a straightforward process yet, as in exploration seismic, because of the size and complex physics of the entire crust and mantle we have to take into account.
Unfortunately,
comparisons with other models will always be rather qualitative, because we have no way of using such models in 3D wave simulations, mainly because 3D mantle models are difficult to combine with 3D crustal models. Underneath the oceans there tends to be a `gap' between the mantle model and the crust, whereas underneath the continents there tends to be a region of overlap, and this makes `gluing' the two together challenging. Thus it may be unfair to assess the models with crustal corrections with 3D numerical wave simulations.}

@@

\rev{I recommend therefore some restructuring of the paper by relating the modeling results to the adjoint tomography approaches. To stay on the topics raised in sections 1--9, I suggest the authors demonstrate with specific examples that the complex GLAD-M workflow is the only viable approach for making progress in global-scale seismic tomography. The paper will become more technical in tone and will cater primarily to the seismological community but that should be acceptable for GJI. Image interpretations in terms of Earth's structure and dynamics could be published in journals with a somewhat broader readership. Restructuring will involve new work for the leading student author, but I think it is worth the effort.
}

\response{We have substantially restructured the manuscript along the lines suggested.}

\rev{1. Recalculation of sensitivity kernels after each iteration is expensive. Is it necessary?
It would be helpful to see examples of how surface wave kernels changed from iteration 0 to 25 for ``simple'' paths (e.g., in ocean basins) and for ``complex'' paths (e.g., along ocean-continent margins). Are there good examples of waveforms that demonstrate that M25 is explaining surface-wave multipathing because kernels are recalculated for a 3-D structure?
}

\response{We do not calculate sensitivity kernels, as in the Dahlen \& Nolet banana-doughnut approach! Instead, we directly calculate contributions to the model gradient, which involves assimilation of the measurements for the current model. Thus, this model update must be recalculated at each iteration.
There are lots of examples of the effects of 3D structure on such kernels in the literature. See, for instance, Tape et al. (2007) for multi-pathing surface waves, or synthetic studies such as Koroni et al. (2017) for the effects of superposition of multiple arrivals. However, thanks for the comment; we want to study such observations in follow-up papers.}

@@

\rev{2. Along the same lines, do the sensitivity kernels for body waves through the deep mantle change from iteration 0 to 25? Always or only when, for example, when they skirt the LLSVP? Would it be sensible to recalculate the high-frequency body-wave kernels only every 5th or 10th iteration to safe computational time? Are there good reasons not to do this?
}

\response{See our response to the previous comment. We do not calculate individual sensitivity kernels a la Dahlen \& Nolet. We compute ``event kernels" for each event combining all body- and surface-wave measurements together. Even if one may argue that lower mantle kernels should be simple, D'' and the upper mantle are definitely not. For instance, multiply reflected body waves from the free surface can show complicated patterns, as one might expect. Furthermore, even though the lower mantle portion might be relatively simple, the crustal and upper-mantle portions of the kernels may still be affected by strong heterogeneities. Overall it does not cost extra for us to run 3D kernels for body-wave phases sampling the lower mantle. 
%\fromebru{I think we can put such a figure as well (I mean in the response, not sure needed for the paper but may go into SOM as well). The kernels must definitely change for paths across the plumes. Also we may be able to give what I encountered in GLAD-M15 before taking full attenuation into account in adjoint simulations as an evidence that we need to compute exact kernels to further decrease the misfit and improve the model}
%\toebru{Going back to Carl and Qinya's papers, there are lots of such examples in the literature. There is no need to belabor this.}
}

@@

\rev{3. I would like to see examples of recorded and synthetic data at long (T > 250 s), and short (T < 10 s) periods to illustrate that M25 is (close to) reproducing waveform segments and data types not yet selected by FLEXWIN. How good is the fit to waveform segments (e.g., ScS) and data types (e.g., normal mode splitting) not included in the inversion?
}

\response{Calculating synthetics at periods less than 17 s is a very expensive proposition! Going to 9~s involves a 16X increase in the computational cost. The long period corner of the period band is set by the Cowling approximation,
which limits simulations to periods up to $\sim250--300$~s. However, all these suggestions make interesting projects for further investigation.
}

@@

\rev{f) Figure 3. This figure is fairly busy and not informative. If the authors decide to keep it, I suggest the oceans and continents are simply white.
}

\response{We like this figure, which gives a good idea of station `hit count'. It has been combined with the source figure.
}

\rev{g) Page 8. I have read the statement on SAC before in Jeroen's papers and I apologize for my quibble. The ASDF format may very well be ideal but there are options other than SAC. Many seismologists, including yours truly 20 years ago, developed data and metadata containers that are self-describing, compact, and ``cat-able'' to improve I/O and processing speeds. The bullets on page 9 are too vague.
}

\response{Point taken.  Indeed other data formats have been developed based on a similar logic. We emphasize ASDF because of its compatibility with HPC systems. We have removed the section on ASDF because the format is already discussed in the original ASDF paper, and in the Bozdag et al.~(2016) paper.}

@@

\rev{h) Page 9. It is not clear to me which part(s) of the seismogram the data categories encompass. Can waveform segments classified as ``17-40 s body waves'' overlap with segments from the ``40-100 s body waves'' group? It would be helpful to show four seismograms for the different frequency bands that indicate where windows are typically chosen.
}

\response{The windows are taken independently of categories, based on the quality of fit. The windows in different categories do not necessarily overlap. Examples of such windows may be found in Bozdag et al.~(2016).}

\rev{i) Page 10. Was the imbalance in M15 (e.g., thousands of paths between circum-Pacific events and the USArray) the motivation for the source and receiver weights in equations 5 and 6?
}

\response{Yes. M15 tried to compensate for this by using kernel density as a preconditioner in the post-processing stage, but the new approach demonstrably leads to faster and better convergence (see Ruan et al.~2019).}

@@

\rev{j) Figure 6. Both PS and SP can be strong on the vertical and radial components. So, change the label ``PS'' to ``SP+PS''. Why are there few segments close to the PPP arrival (beyond 100 degrees)? Since it is used in the caption, I would use seconds along the y-axis (also in the next two figures).
}

\response{Points taken. PPP is not prominent, and picked rarely by Pyflex with conservative selection criteria to avoid bad-quality windows. In addition, to linearize our inversions, we do not select any window with large time shifts or amplitude anomalies until observed and synthetic waveforms become comparable. It is ongoing research to improve automated phase picking algorithms.}

@@

\rev{l) Table 1. Clarify ``changes in fit'' or maybe show residual misfit instead which is used frequently. Is it correct that the 40-100 s period surface wave fit has improved the most? This would indicate that the 10 iterations have primarily adjusted the uppermost mantle.
Is the misfit reduction smallest for the ``overtones surface waves'' (the S multiples arriving before the fundamental mode) (i.e., 40-100 s body waves)?
This might indicate that the mantle transition is still resolved poorly (or was already resolved very well by S362ANI).
Section 8.4 could use more discussion regarding the misfits of the various categories and what it implies for the resolution of dVs (and dVp) as a function of depth.
}

\response{This is the variance reduction for the subset of measurements in each of the 12 categories. Surface waves show more of a reduction because the waveforms are more complex, and moving them in phase gives a greater variance reduction than improving the fit to a simpler body waveform. The categories reflect pass bands and components, not depth resolution..}

\rev{m) Figure 10. The histograms of Figure 10 also suggest the modeling of the 40-100 s surface waves has been most successful. Maybe the overtones should have more weight in the next iterations?
}

\response{See the earlier explanation.
Surface wave fits are much worse than body wave fits prior to inversion, because the former waveforms are dispersed.
Thus there are more gains to be had. We don't strip overtones, so these are not fit individually, nor can they be weighted separately.
Instead, we fit full surface waveforms. However, we are working on new measurement techniques where higher modes and any scattered waves with smaller amplitudes can be better taken into account than cross-correlation measurements. See Bozdag et al.~(2011) and the recent paper by Yuan et al.~(2020) for such measurements, which we aim to incorporate in our future iterations. 
}

@@

\rev{p) Page 26 and Figure 16. The authors must include panels for M15 in Figure 16 and discuss what has changed.
Given the variance reductions I suspect most adjustments have been made in the upper mantle. }

\response{We have now added comparison of M25 to M15 in the revised manuscript and in the SOM.}

\rev{(i) Although the authors invert for dVp and dVs, the resolution of dVp is poorer than of dVs, particularly in the uppermost mantle (constrained by surface waves) and in D'' (constrained by diffracted waves).}

\response{We are not sure if this is a general comment for seismic tomography or for our models specifically. S-wave perturbations are in general larger than those of P waves. It is also likely that we have less P-wave measurements than S-wave measurements because of the smaller amplitudes of P waves. Therefore P-wave evolution can be slower than that of S waves. The smoother P-wave model of GLAD-M25 compared to other P-wave models is also likely due to our minimum period of body waves, which is 17~s.}

@@

\rev{(ii) The strength of structure in the models is for a large part determined by regularization. As shown in Ritsema et al. 2011 we can change the amplitude of S40 (and the width of the resolution kernels) by changing the damping without compromising the data fit significantly. Comparing amplitudes of structures from different models without considering resolution differences is a tricky thing.}

\response{We agree that the amplitude of classical tomographic models heavily depend on subjective decisions of damping and other factors while it is required for the interpretation of the thermochemical structure and dynamics of the mantle. Unfortunately we cannot make a reliable assessment of amplitudes. All we can say is that our results show significant increases in amplitudes compared to M15 specifically for plume and slab structures where our non-linear gradient-based optimization does not involve a damping factor.}

@@

\rev{q) In Figure 18 I suggest to include the model by Krischer et al. (JGR; 2018). Model US22 is perhaps not that useful in the context of this comparison since it is based on a similar modeling approach and data set as M25.}

\response{We could use Krischer et al. (2018), however it looks like it is smoother than US22. Our point is that we have started achieving the resolution of continental-scale studies in global inversions wherever we have dense coverage.}

\rev{r) Since SL2013NA (a ``ray'' model) doesn't appear too different from M25 it is not clear from the visual comparison) whether modeling improvements (i.e., better theories) matter.
(Note that I suspect they do matter for surface waves because your variance reduction is large for the surface wave data -- see my point 1).}

\response{Not sure if they are very similar, in fact they show strong discrepancies below 100~km. }

@@

\rev{s) The differences among the models at depths larger than 300 km in Figure 18 is striking.
The authors draw the reader's attention to anomalies with well-understood tectonic origins but at face value the images are really very different especially for US22 (and M25) and SL2013NA.
The latter model is based on a large set of overtone surface wave data that should sample the MTZ very well. The misfit reduction obtained for the 40-100 s body + surface waves (presumably waveform segments after SS and before the fundamental mode surface wave) is relatively poor, so perhaps the MTZ in M25 is still poorly resolved (or S362 was an excellent model to begin with).
At any rate, this need further discussion. Transition zone resolution is still a sticky point and is important for understand convective flow.}

\response{Due to our cross-correlation measurements higher-modes may well be underrepresented in the current inversions. We are working on other misfit functions to effectively incorporate smaller-amplitude signals (e.g., Bozdag et al. 2011, Yuan et al. 2020) together with better window-selection strategies.}

\rev{t) Figure 19. Perhaps include SL2013sv and the model by Rickers et al. (EPSL; 2013).
}
\response{The Rickers et al. model is more focused on the upper mantle, thus we chose to compare the Icelandic plume to those in other global models to see the continuation of the plume. However, we now discuss it in the revised manuscript.}

### 2020-02-19 main.tex

Another way to evaluate model quality is by assessing the distribution
of measurements in the various categories.
Again we use all 1,480 events and a set of identical windows on all three components
to assess GLAD-M25, GLAD-M15, and S362ANI+CRUST2.0.
The total number of selected windows,
determined using model GLAD-M25, exceeds 18~million.
Fig.~\ref{fig:phase_hist} shows histograms of the resulting phase
measurements in all twelve measurement categories.
We observe that the distributions generally become better centered on zero,
and that all standard deviations are steadily reduced.
The inversion is aimed at reducing the overall misfit,
so there are small tradeoffs between different measurement categories.
%\cyan{How were source corrections applied to GLAD-M15 and S362ANI+Crust2.0? Did you apply the correction terms computed for GLAD-M25 to all models?} I believe so.

@@

First, we repeated the misfit assessment discussed in
Section~\ref{section:Misfit assessment} for the held-out database.
Table~\ref{table:misfit_reduction_M15_M25_360} summarizes the reductions
in misfit for the held-out database for model GLAD-M25 compared to GLAD-M15
as well a S362ANI+CRUST2.0.
It is encouraging to see that the reductions are comparable to the
values tabulated in Table~\ref{table:misfit_reduction_M15_M25} for the actual inversion.
Overall, the reductions are slightly smaller, partly due to the fact that we did not
conduct full CMT inversions for the held-out database.
%\cyan{not sure..}

@@

Vertical cross sections are plotted relative to the spherical average of GLAD-M25,
as discussed in Appendix~\ref{sec:1Dmodel}.
For plotting purposes and use in other applications,
an expansion of the model in spherical harmonics and radial B-splines is
discussed in Appendix~\ref{sec:shanalysis}
%\cyan{but the models were not expanded to spherics harmonics to plot in this paper, right?}. No.

@@

As illustrated in Fig.~\ref{fig:M25-AFAR},
we see an enhancement of the Afar plume relative to GLAD-M15.
The 3D morphology of the African superplume shows a mantle plume rising from the core-mantle boundary with a broad base tilting northward
%\cyan{do you know why the African plume tilts? I think Pacific plume tilts likely because of the plate motion. GLAD-M25 generally shows that this tilt is happening widely in the upper mantle and perhaps the top of the lower mantle. But the African plume's tilt start from the CMB}  No. It's in the Texas model too.
with a thinning neck, before it reaches the 660~km phase boundary and flattens.
It then develops into two sub-plumes ---Afar and East Lake Victoria--- supplying the East Africa rift.
The development of the superplume and its 3D morphology might be associated with the thermal and chemical states in the lower mantle and volatiles in the mantle transition zone, a possibility we will explore in a future study.

% Subduction along the Andes in South America is imaged from south to north all the way down to the transition zone and beyond in a more continuous way compared to GLAD-M15, as shown in Fig.~\ref{fig:S_America}.
% In the contour plot of positive shear-wavespeed perturbations we observe three gaps, which may correspond to ridges or volcanic plateaus on the subducted Nazca plate, and indicate a weakened or even torn slab \citep[e.g.,][]{Hu2016} 
%\cyan{I think it is a great story but would you like to put the contour plot here or reserve for the future paper?}. OK.

@@

Overall, the three models are in reasonably good agreement.
At shallow depths, model GLAD-M25 shows more distinct continental crust and cratons, thanks to the inclusion of surface waves.
But in the upper and mid mantle,
the models share similar attributes.
%Relatively smoother slabs in GLAD-M25 are likely related to our choice of smoothing considering the minimum period of body waves in our simulations, $\sim17$~s, which is longer than typical teleseismic body-wave tomography studies.

>>>>

Another way to evaluate model quality is by assessing the distribution
of measurements in the various categories.
Again we use all 1,480 events and a set of identical windows on all three components
to assess GLAD-M25, GLAD-M15, and S362ANI+CRUST2.0.
The total number of selected windows,
determined using model GLAD-M25, exceeds 18~million.
Fig.~\ref{fig:phase_hist} shows histograms of the resulting phase
measurements in all twelve measurement categories.
We observe that the distributions generally become better centered on zero,
and that all standard deviations are steadily reduced.
The inversion is aimed at reducing the overall misfit,
so there are small tradeoffs between different measurement categories.

@@

First, we repeated the misfit assessment discussed in
Section~\ref{section:Misfit assessment} for the held-out database.
Table~\ref{table:misfit_reduction_M15_M25_360} summarizes the reductions
in misfit for the held-out database for model GLAD-M25 compared to GLAD-M15
as well a S362ANI+CRUST2.0.
It is encouraging to see that the reductions are comparable to the
values tabulated in Table~\ref{table:misfit_reduction_M15_M25} for the actual inversion.
Overall, the reductions are slightly smaller, partly due to the fact that we did not
conduct full CMT inversions for the held-out database.

@@

Vertical cross sections are plotted relative to the spherical average of GLAD-M25,
as discussed in Appendix~\ref{sec:1Dmodel}.
For plotting purposes and use in other applications,
an expansion of the model in spherical harmonics and radial B-splines is
discussed in Appendix~\ref{sec:shanalysis}

@@

As illustrated in Fig.~\ref{fig:M25-AFAR},
we see an enhancement of the Afar plume relative to GLAD-M15.
The 3D morphology of the African superplume shows a mantle plume rising from the core-mantle boundary with a broad base tilting northward
with a thinning neck, before it reaches the 660~km phase boundary and flattens.
It then develops into two sub-plumes ---Afar and East Lake Victoria--- supplying the East Africa rift.
The development of the superplume and its 3D morphology might be associated with the thermal and chemical states in the lower mantle and volatiles in the mantle transition zone, a possibility we will explore in a future study.

@@

Overall, the three models are in reasonably good agreement.
At shallow depths, model GLAD-M25 shows more distinct continental crust and cratons, thanks to the inclusion of surface waves.
But in the upper and mid mantle,
the models share similar attributes.
